A Survey of Possible Exoskeleton Control Architectures and Algorithms\\ (Draft 3.0)

### 2015-09-25 control.tex
cga updates

%\section{Symbiosis and Autonomy, Handling Errors, and Superhuman Reflexes}
A later white paper will discuss
possible combinations of response to user and exoskeleton
autonomy including balance control; response to trips, slips,
stumbles, and fumbles; task assistance and guidance (guide operator to
doorknob, button, or light switch); superhuman response to external
perturbations such as projectiles and explosions; and autonomous
execution (operator/exoskeleton symbiosis with multi-tasking)?

A later white paper will discuss how to handle malfunctions and damage to the
system. Basically this boils down to fault detection, and switchover to
an appropriate safe mode of operation.

A later white paper will evaluate using simulation
the various control approaches' robustness to modeling error and unmodeled
dynamics.

@@

Low level control presents an API (Application Programmer Interface)
for higher control levels
to access the actuators, sensors, and other robot hardware~\cite{IEEE07271219}.
It implements high servo rate processing including feedback control and online
optimization.
This level provides an idealization of the exoskeleton in two ways: 
What degrees of freedom (DOF) are controlled, and how is a degree of
freedom controlled?

@@

It is useful to provide human programmers
with idealized abstractions of control of a single degree of
freedom. Here are some common alternatives that could be provided by low level control:
\begin{itemize}
\item
Position control. The joint moves to a specified position and stays there independent
of load.
\item
Velocity control. The joint moves at a specified velocity independent of load.
\item
Force or torque control. The joint provides a specified force or torque independent
of its motion.
\item
Impedance control: The user specifies a desired position (for a spring), and commands
a desired
stiffness, damping, and potentially a modified inertia or moment of inertia.
Modifying inertia is more dangerous than providing variable active stiffness
or damping.
Or (almost equivalently),
the user specifies a desired force or torque
which changes according to the position, velocity,
and acceleration of the joint.
\item
Valve control (for fluidic systems):
\begin{equation}
\vu = \mK_{\vq} (\vq - \vq_d) + \mK_{\dot{\vq}} ( \dot{\vq} - \dot{\vq}_d )
+ \mK_{\vtau} ( \vtau - \vtau_d )
+ \mK_{\dot{\vtau}} ( \dot{\vtau} - \dot{\vtau}_d ) + \vu_{\ff}
\end{equation}
where $\vu$ is a valve command vector (with a valve command for each actuator),
$\vq$ is the vector of joint angles, $\dot{\vq}$ are joint velocities,
$\vtau$ are actutor forces or joint torques, and $\dot{\vtau}$ is the rate of change
of the forces or torques (the time derivative of $\vtau$). 
The subscript $d$ indicates a desired value,
$\mK_{\vq}$ is a joint position gain matrix,
$\mK_{\dot{\vq}}$ is a joint velocity gain matrix, 
$\mK_{\vtau}$ is a joint force/torque gain matrix,
and $\mK_{\dot{\vtau}}$ is a joint force/torque derivative gain matrix.
$\vu_{\ff}$ are feedforward valve commands.
This formulation takes into account the most important hydraulic actuator dynamics,
and allows full state feedback
(including $\vtau$ as part of the state), pole placement, and linear quadratic
regulator (LQR) control of
a third order model of actuator plus rigid body dynamics.
If electric motor and power amplifier dynamics are taken into account in electric
motor systems, there would be an analogous equation for the commanded
motor voltage of
an analog power amplifier or
the control input of a PWM amplifier.
The $\mK_{\vtau} ( \vtau - \vtau_d )$ could be replaced or augmented
by a servo on piston differential pressure in a fluidic system or
a current servo in an electrical system.
\item
Online optimization of accelerations, actuator forces and torques, and contact forces
using quadratic programming. See Section \ref{sec:qp} for further explanation.
\end{itemize}

@@

Exoskeletons are similar to robots, but there are several important differences.
There are two sets of contact forces: operator-exoskeleton (which we denote
with a subscript $\ox$ as in $\vf_{\ox}$) and exoskeleton-world (which we denote
with a subscript $\xw$ as in $\vf_{\xw}$). If the operator is tightly strapped or
rigidly held by other types of physical constraints, we can lump the operator
body parts with the exoskeleton parts which they are attached to in terms
of link inertias, locations of center of mass, and moments of inertia,
and just think of the whole system as one robot (we indicate this with
the subscript $\ox$ on inertial, Coriolis and centripetal, and gravitational (ICCG)
terms in the dynamics equations).
We can also separate out the operator dynamics (indicated by
the subscript $\myo$ on ICCG terms) from the exoskeleton dynamics (indicated by
the subscript $\myx$ on ICCG terms) if this is more convenient
(we separate them out and ignore operator dynamics in what immediately follows).
If the operator is loosely connected to the exoskeleton, this introduces a
huge modeling and control problem, and would also be very dangerous for the
operator (as riding a bull or bucking horse in a rodeo is dangerous to a rider).
We do not address the loosely connected case in this paper.

The dynamics of an operator standing on the ground (or on parts of the exoskeleton
that are on the ground):
\begin{equation}
\mM_{\myo}(\vq) \ddot{\vq} + \mC_{\myo}(\vq,\dot{\vq}) + \mG_{\myo}(\vq)
+ \mJ_{\ow}^{\tr}(\vq) \vf_{\ow} = \vtau_{\myo} - \mJ_{\ox}^{\tr}(\vq) \vf_{\ox}
\label{eq:odynamics}
\end{equation}
where $\vq$ are the operator joint angles,
$\dot{\vq}$ and $\ddot{\vq}$ are the corresponding
joint velocities and accelerations,
$\mM_{\myo}$ is the operator inertia matrix, 
$\mC_{\myo}()$ are the operator Coriolis and
centripetal forces, 
$\mG_{\myo}()$ are the operator gravitational forces,
$\mJ_{\ow}^{\tr}(\vq) \vf_{\ow}$ (Jacobian matrix multiplied with the
contact forces) are the operator-world contact forces 
expressed as exoskeleton joint torques, 
$\vtau$ are the operator joint torques due to muscle activity and passive tissue
mechanics,
and 
$\mJ_{\ox}^{\tr}(\vq) \vf_{\ox}$
are the operator-exoskeleton forces expressed
as operator joint torques.
We assume the operator joint friction is negligable.

@@

To understand these equations better, it is useful to have all Jacobians equal
to the identity matrix. This means the operator is directly applying torques
at the exoskeleton joints ($\vtau_{\ox}$) (a realistic assumption), 
and so is the world ($\vtau_{\ow}$ and $\vtau_{\xw}$)
(a very unrealistic assumption). In this case,
the operator dynamics are:
\begin{equation}
\mM_{\myo}(\vq) \ddot{\vq} + \mC_{\myo}(\vq,\dot{\vq}) + \mG_{\myo}(\vq)
+ \vtau_{\ow} = \vtau_{\myo} - \vtau_{\ox}
\end{equation}
the exoskeleton dynamics are:
\begin{equation}
\mM_{\myx}(\vq) \ddot{\vq} + \mC_{\myx}(\vq,\dot{\vq}) + \mG_{\myx}(\vq)
+ \mF_{\myx}(\vq,\dot{\vq})
+ \vtau_{\xw} = \vtau_{\myx} + \vtau_{\ox}
\end{equation}
and the combined dynamics are:
\begin{equation}
\mM_{\ox}(\vq) \ddot{\vq} + \mC_{\ox}(\vq,\dot{\vq}) + \mG_{\ox}(\vq)
+ \mF_{\myx}(\vq,\dot{\vq})
+ \vtau_{\ow}
+ \vtau_{\xw} 
= \vtau_{\myo} + \vtau_{\myx}
\label{eq:oxdynamics}
\end{equation}

@@

One control option is to use only passive controls on a joint or on
all joints, and no active controls
or actuators.
Such a system would rely on counterbalancing, mechanical springs, air springs, etc.
to reduce forces. 
An operator would rely on mechanical backdrivability to drive
exokeleton to do tasks.
These passive devices have no controls, and can be included in the dynamics equations.
In this case, 
$\vtau_{\myx}$ is zero for any passive degree of freedom.

@@

@@

Active exoskeleton gravity compensation is similar 
to counterbalancing or physical gravity compensation, except it is done
with exoskeleton actuators to cancel the exoskeleton weight (but not inertia):
\begin{equation}
\vtau_{\myx} = \hat{\mG}_{\myx}(\hat{\vq})
\end{equation}
where $\vtau_{\myx}$ are the exoskeleton joint torques, 
$\hat{\mG}_{\myx}()$ are the exoskeleton
gravitational forces calculated using an imperfect model, 
and $\hat{\vq}$ is the estimated configuration of the operator (and exoskeleton),
The operator relies on mechanical backdrivability to drive
the exokeleton to do tasks.
The weight of the operator can also be supported by the exoskeleton in
full gravity compensation:
\begin{equation}
\vtau_{\myx} = \hat{\mG}_{\ox}(\hat{\vq})
\end{equation}
where $\hat{\mG}_{\ox}()$ are the combined operator and exoskeleton
gravitational forces calculated using an imperfect model,

@@

Friction can also be compensated.
\begin{equation}
\vtau_{\myx} = \hat{\mF}_{\myx}(\hat{\vq},\hat{\dot{\vq}}) 
\end{equation}
are the estimated friction torques at the estimated exoskeleton configuration
and velocity.
Note that these may be load dependent, and may require additional state to
correctly handle stiction and hysteresis effects.

@@

What is admittance?
\url{https://en.wikipedia.org/wiki/Admittance}
Compliance, inverse damping, inverse mass are components of an admittance,
as they map force and torques to kinematic variables (position, velocity, and
acceleration).

We can implement operator force control 
if measurements of the operator contact forces with
the exoskeleton are available.
The exoskeleton moves as to create as little force on the operator as possible 
assuming the actuators are torque sources:
\begin{equation}
\vtau_{\myx} = \vtau_{\xinvdynw} - \mJ_{\ox}^{\tr}(\vq) {\vf_{\ox}}_d
- \mJ_{\ox}^{\tr}(\vq) \mK_{\vf_{\ox}} ( \hat{\vf}_{\ox} - {\vf_{\ox}}_d )
\end{equation}
$\mK_{\vf_{\ox}}$ is a gain matrix, $\mJ_{\ox}$ is the Jacobian matrix for the operator
force $\vf_{\ox}$
Note that this does not require inverting the Jacobian matrix.
We will see that the position control version does invert the Jacobian matrix.

@@

So far we have ignored or canceled the effect of 
the exoskeleton contact with the outside world.
A different control objective could be to map the external forces to 
desired operator forces:
\begin{equation}
{\vf_{\ox}}_d = F_{\ox-\xw} ( \vf_{\xw} )
\end{equation}
and cancel the external forces not transferred in the inverse dynamics.
Everything in the above equation is estimated, so we are dropping the hats 
$\hat{\mbox{}}$.

@@

This is usually done by using stored trajectories or trajectories generated
online that specify desired position $\vq_d$, 
desired velocity $\dot{\vq}_d$, and
desired acceleration $\ddot{\vq}_d$.
The feedforward method is to compute the needed torques (potentially offline)
using inverse
dynamics (including the operator link inertias, link centers of mass, and link
moments of inertia in the dynamic model) based on the desired postion, 
desired velocity, and desired acceleration. Trajectory tracking errors are compensated
for by feedback control designed using some other method. In most robotics
fixed gain independent joint linear control is used.
This method is safest in the sense that the feedback controller can be tested
and verified independently of the desired trajectories and dynamic models.

@@

Another control objective would be to allow the exoskeleton-world contact to
be compliant or have an assigned impedance.
Let's describe a small position offset of the exoskeleton-world
contact points as $\vq_w$. We know that:
\begin{equation}
\dot{\vq}_w = \mJ_{\xw}(\vq) \dot{\vq}
\end{equation}
and that
\begin{equation}
\vtau_{\myx} = \mJ_{\xw}^{\tr}(\vq) \vf_{\xw} 
\end{equation}
This allows us to relate stiffness and damping in the external world coordinates
with the stiffness and damping in exoskeleton joint coordinates.
Here is the relationship for damping:
\begin{equation}
\vtau_{\myx} = \mJ_{\xw}^{\tr}(\vq) \mK_{\dot{w}} \mJ_{\xw}(\vq) \dot{\vq}
\end{equation}
So the damping in joint coordinates is $\vtau_{\myx} = \mK_{\dot{\vq}} \dot{\vq}$ and
therefore
\begin{equation}
\mK_{\dot{\vq}}(\vq) = \mJ_{\xw}^{\tr}(\vq) \mK_{\dot{w}} \mJ_{\xw}(\vq)
\end{equation}
The mapping is the same for stiffnes:
\begin{equation}
\mK_{\vq}(\vq) = \mJ_{\xw}^{\tr}(\vq) \mK_{w} \mJ_{\xw}(\vq)
\end{equation}
So we should assign the exoskeleton joint position gains to be $\mK_{\vq}(\vq)$ and
the joint velocity gains to be $\mK_{\dot{\vq}}(\vq)$ in equation~\ref{eq:impedance}.

@@

This section describes how to handle underactuation and overactuation using
``floating body dynamics.''
We have been sweeping this
issue under the rug so far. It is now useful to be more explicit.
There are two kinds of underactuation we want to consider. The first
is ``floating body dynamics'', where the robot is not bolted to the floor.
Contacts are made and broken during walking, running, and many other behaviors,
and contacts slip.
Humanoid robots (and humans) have floating body dynamics.
A ``root'' location (usually somewhere
on the pelvis) is chosen, and the position, velocity, orientation, and angular
velocity of that point and the rigid body it is on
are included in the state vector $\vq$. The corresponding dimensions of the
velocity vector $\dot{\vq}$ are not actuated. A Jacobian matrix
$\mJ_{\myx}$ is used to map from actuation variables $\vu$ to joint torques $\vtau$:
\begin{equation}
\vtau_{\myx} = \mJ_{\myx} \vu_{\myx}
\end{equation}
The rows in $\mS$ corresponding to the root linear and angular velocities
are all zero.
To maintain compatability in our joint representation, we will similarly
map operator ``muscle commands'' $\myo$ to operator joint torques $\vtau_{\myo}$.
\begin{equation}
\vtau_{\myo} = \mJ_{\myo} \vu_{\myo}
\end{equation}

The second kind of underactuation appears in exoskeletons where we rely on the
operator to actuate a particular joint. For example, many lower body
exoskeletons only actuate the joints in the sagittal plane, leaving lateral
and twist (yaw) joints unactuated or passively actuated. 
Many ``exoskeletons'' or orthoses only actuate one joint on each limb (typically
the ankle or knee). In these situations the question arises as to what is
the scope of control: should the device only consider
just the actuated joints,
or should it consider a wider range of operator joints? Many of these devices
do not have enough sensing (such as an IMU) to even estimate the state of other
operator joints. We believe that for high performance the entire state of the
operator must be considered in control, and we describe a way to do it in
Section~\ref{sec:qp}.
However, we are not aware of any system that goes this far in terms of control.

@@

``Missing joint''
underactuation can be handled by reducing the size of the $\vu_{\myx}$ vector
and changing the matrix $\mJ_{\myx}$ to reflect the missing actuation.
We can have redundant actuation (more than one actuator across any particular
joint). Here we augment $\vu_{\myx}$ and $\mJ_{\myx}$ to accomodate extra actuators.
We can have multijoint actuation, which can be handled in $\vu_{\myx}$ 
and $\mJ_{\myx}$ as well.

@@

{\bf Open loop control} continues to work with underactuation
as long as a drive function can be found
to generate the desired behavior.

@@

Full (operator) body
{\bf online inverse dynamics} as described in Section~\ref{sec:invdyn}
no longer works if $\mJ_{\myx}$ is not invertable (not square or not full rank).
In the underactuated and overactuated cases $\mJ_{\myx}$ is not square. 
For a partial (operator) body
exoskeleton we can rewrite the dynamics in terms of only those
joints the exoskeleton actuates, and invert those dynamics to cancel exoskeleton
effects on those joint. This is what has been done with the BLEEX exoskeleton,
for example.~\cite{}.

@@

Quadratic programming~\cite{Wikipedia} solves the following problem:
Minimize
\begin{equation}
\vx^{\tr} \mQ \vx + c^{\tr} \vx
\end{equation}
with inequality constraints
\begin{equation}
\mA_1 \vx \le \vb_1
\end{equation}
and equality constraints
\begin{equation}
\mA_2 \vx = \vb_2
\end{equation}
The optimization variable $\vx$ includes accelerations $\ddot{\vq}$, actuator
commands $\vu$, and contact forces $\vf$.
For robots the discrepancy between actual and desired accelerations, actuator command
magnitude, and contact force magnitude are all minimized:
\begin{equation}
C_r(\vx) = (\vq - \vq_d)^{\tr} \mW_{\vq} (\vq - \vq_d) + \vtau^{\tr} \mW_{\vtau} \vtau
+ \vf^{\tr} \mW_{\vf} \vf
\end{equation}
$\mW_{\vq}$, $\mW_{\vtau}$, and $\mW_{\vf}$ are weight matrices to trade off the
different errors/magnitudes and error/magnitude directions.
The dynamics (equation~\ref{eq:dynamics2}) are enforced as an equality constraint.
Actuator limits, friction limits, center of pressure location constraints,
and unidirectional contact constraints are all
enforced as inequality constraints.
Variants of this optimization
were performed every 1-2ms by various teams to control the
Atlas robots in the DARPA Robotics Challenge.

This approach can be used to achieve desired joint or synergy (such as center
of mass) accelerations, center of mass dynamic behavior, center of pressure location,
weight distribution, maintenance of a desired body part position, orientation,
velocity, or trajectory. and joint/actuator position and velocity limits (indirectly).

\subsection{Controlling operator forces and perceived exoskeleton impedance}

>>>>

%\section{Symbiosis and Autonomy, Handling Errors, and Superhuman Reflexes}
A later white paper could discuss
possible combinations of response to user and exoskeleton
autonomy including balance control; response to trips, slips,
stumbles, and fumbles; task assistance and guidance (guide operator to
doorknob, button, or light switch); superhuman response to external
perturbations such as projectiles and explosions; and autonomous
execution (operator/exoskeleton symbiosis with multi-tasking)?

A later white paper could discuss how to handle exoskeleton
malfunctions and damage to the
system. Basically this boils down to fault detection, and switch over to
an appropriate safe mode of operation.

A later white paper could evaluate using simulation
the various control approaches' robustness to modeling error and unmodeled
dynamics.

@@

Low level control presents an API (Application Programmer Interface)
for higher control levels
to access the actuators, sensors, and other robot hardware~\cite{IEEE07271219}.
It implements high servo rate processing including feedback control and online
optimization.
This level provides an idealization of the exoskeleton in two ways: 
{\it What degrees of freedom (DOF) are controlled?,} and 
{\it How is a degree of freedom controlled?}

@@

It is useful to provide human programmers
with idealized abstractions of control of a single degree of
freedom. Here are some common alternatives that could be provided by low level control:
\begin{itemize}
\item
Position control. The joint moves to a specified position and stays there independent
of load.
\item
Velocity control. The joint moves at a specified velocity independent of load.
\item
Force or torque control. The joint provides a specified force or torque independent
of its motion.
\item
Impedance control: The user specifies a desired position (for a spring), and commands
a desired
stiffness, damping, and potentially a modified inertia or moment of inertia.
Modifying inertia is more dangerous than providing variable active stiffness
or damping.
Or (almost equivalently),
the user specifies a desired force or torque
which changes according to the position, velocity,
and acceleration of the joint.
\item
Valve control (for fluidic systems):
\begin{equation}
\vu = \mK_{\vq} (\vq - \vq_d) + \mK_{\dot{\vq}} ( \dot{\vq} - \dot{\vq}_d )
+ \mK_{\vtau} ( \vtau - \vtau_d )
+ \mK_{\dot{\vtau}} ( \dot{\vtau} - \dot{\vtau}_d ) + \vu_{\ff}
\end{equation}
where $\vu$ is a valve command vector (with a valve command for each actuator),
$\vq$ is the vector of joint angles, $\dot{\vq}$ are joint velocities,
$\vtau$ are actuator forces or joint torques, and $\dot{\vtau}$ is the rate of change
of the forces or torques (the time derivative of $\vtau$). 
The subscript $d$ indicates a desired value,
$\mK_{\vq}$ is a joint position gain matrix,
$\mK_{\dot{\vq}}$ is a joint velocity gain matrix, 
$\mK_{\vtau}$ is a joint force/torque gain matrix,
and $\mK_{\dot{\vtau}}$ is a joint force/torque derivative gain matrix.
$\vu_{\ff}$ are feedforward valve commands.
This formulation takes into account the most important hydraulic actuator dynamics,
and allows full state feedback
(including $\vtau$ as part of the state), pole placement, and linear quadratic
regulator (LQR) control of
a third order model of actuator plus rigid body dynamics.
If electric motor and power amplifier dynamics are taken into account in electric
motor systems, there would be an analogous equation for the commanded
motor voltage of
an analog power amplifier or
the control input of a PWM amplifier.
The torque error term $\mK_{\vtau} ( \vtau - \vtau_d )$ could be replaced or augmented
by a servo on piston differential pressure in a fluidic system or
an electrical current servo in an electric system.
\item
Online optimization of accelerations, actuator forces and torques, and contact forces
using quadratic programming. See Section \ref{sec:qp} for further explanation.
\end{itemize}

@@

Exoskeletons are similar to robots, but there are several important differences.
There are two sets of contact forces: operator-exoskeleton (which we denote
with a subscript ${\mbox {ox}}$ as in $\vf_{\ox}$) and exoskeleton-world (which we denote
with a subscript ${\mbox {xw}}$ as in $\vf_{\xw}$). If the operator is tightly strapped or
rigidly held by other types of physical constraints, we can lump the operator
body parts with the exoskeleton parts which they are attached to in terms
of link inertias, locations of center of mass, and moments of inertia,
and just think of the whole system as one robot (we indicate this with
the subscript ${\mbox {ox}}$ on inertial, Coriolis and centripetal, and gravitational (ICCG)
terms in the dynamics equations as in $\mM_{\ox}$).
We can also separate out the operator dynamics (indicated by
the subscript ${\mbox {o}}$ on ICCG terms as in $\mM_{\myo}$) from the exoskeleton dynamics (indicated by
the subscript ${\mbox {x}}$ on ICCG terms as in $\mM_{\myx}$) if this is more convenient
(we separate them out and ignore operator dynamics in what immediately follows).
If the operator is loosely connected to the exoskeleton, this introduces a
huge modeling and control problem, and would also be very dangerous for the
operator (as riding a bull or bucking horse in a rodeo is dangerous to a rider).
We do not address the loosely connected case in this paper.

The dynamics of an operator standing on the ground (or on parts of the exoskeleton
that are on the ground):
\begin{equation}
\mM_{\myo}(\vq) \ddot{\vq} + \mC_{\myo}(\vq,\dot{\vq}) + \mG_{\myo}(\vq)
+ \mJ_{\ow}^{\tr}(\vq) \vf_{\ow} = \vtau_{\myo} - \mJ_{\ox}^{\tr}(\vq) \vf_{\ox}
\label{eq:odynamics}
\end{equation}
where $\vq$ are the operator joint angles,
$\dot{\vq}$ and $\ddot{\vq}$ are the corresponding
joint velocities and accelerations,
$\mM_{\myo}$ is the operator inertia matrix, 
$\mC_{\myo}()$ are the operator Coriolis and
centripetal forces, 
$\mG_{\myo}()$ are the operator gravitational forces,
$\mJ_{\ow}^{\tr}(\vq) \vf_{\ow}$ (Jacobian matrix multiplied with the
contact forces) are the operator-world contact forces 
expressed as exoskeleton joint torques, 
$\vtau$ are the operator joint torques due to muscle activity and passive tissue
mechanics,
and 
$\mJ_{\ox}^{\tr}(\vq) \vf_{\ox}$
are the operator-exoskeleton forces expressed
as operator joint torques.
We assume the operator joint friction is negligible.

@@

To understand these equations better, it is useful to have all Jacobians equal
to the identity matrix. This means the operator is directly applying torques
at the exoskeleton joints ($\vtau_{\ox}$) (a realistic assumption), 
and so is the world ($\vtau_{\ow}$ and $\vtau_{\xw}$)
(a very unrealistic assumption). In this case,
the operator dynamics are:
\begin{equation}
\mM_{\myo}(\vq) \ddot{\vq} + \mC_{\myo}(\vq,\dot{\vq}) + \mG_{\myo}(\vq)
+ \vtau_{\ow} = \vtau_{\myo} - \vtau_{\ox}
\end{equation}
the exoskeleton dynamics are:
\begin{equation}
\mM_{\myx}(\vq) \ddot{\vq} + \mC_{\myx}(\vq,\dot{\vq}) + \mG_{\myx}(\vq)
+ \mF_{\myx}(\vq,\dot{\vq})
+ \vtau_{\xw} = \vtau_{\myx} + \vtau_{\ox}
\end{equation}
and the combined dynamics are:
\begin{equation}
\mM_{\ox}(\vq) \ddot{\vq} + \mC_{\ox}(\vq,\dot{\vq}) + \mG_{\ox}(\vq)
+ \mF_{\myx}(\vq,\dot{\vq})
+ \vtau_{\ow}
+ \vtau_{\xw} 
= \vtau_{\myo} + \vtau_{\myx}
\end{equation}

@@

One control option is to use only passive controls on a joint or on
all joints, and no active controls
or actuators.
Such a system would rely on counterbalancing, mechanical springs, air springs, etc.
to reduce operator forces. 
An operator would rely on mechanical backdrivability to drive
exoskeleton to do tasks.
These passive devices have no controls, and can be included in the dynamics equations.
In this case, 
$\vtau_{\myx}$ is zero for any passive degree of freedom.

@@

\section{Fixed feedforward control with feedback control}

We can add our favorite feedback control scheme to open loop control.
Now the control is closed loop, but the feedforward control to achieve
a trajectory is designed and executed independently of any feedback control.
In most practical robotics manually designed
fixed gain independent joint linear (PID) feedback control is used.
As previously mentioned, more complex feedback control approaches such as
approaches that coordinate across joints
(full state feedback,
pole placement, linear quadratic regulator (LQR) design, linear quadratic
Gaussian (LQG) design, and linear quadratic Gaussian loop transfer recovery
(LQG-LTR) design)
or nonlinear alternatives that handle discontinuous control problems such as
sliding mode control can also be used.
Feedback gains can be ``gain scheduled'', which means that feedback gains are
pre-tabulated and looked up and deployed
based on the current position, velocity, or some other
quantity (temperature, for example).

@@

Now we will examine feedforward control that varies and is computed based on
the current state of the exoskeleton.
Active exoskeleton gravity compensation is similar 
to counterbalancing or physical gravity compensation, except it is done
with exoskeleton actuators to cancel the exoskeleton weight (but not inertia):
\begin{equation}
\vtau_{\myx} = \hat{\mG}_{\myx}(\hat{\vq})
\end{equation}
where $\vtau_{\myx}$ are the exoskeleton joint torques, 
$\hat{\mG}_{\myx}()$ are the exoskeleton
gravitational forces calculated using an imperfect model, 
and $\hat{\vq}$ is the estimated configuration of the operator (and exoskeleton),
The operator relies on mechanical backdrivability to drive
the exoskeleton to do tasks.
The weight of the operator can also be supported by the exoskeleton in
full gravity compensation:
\begin{equation}
\vtau_{\myx} = \hat{\mG}_{\ox}(\hat{\vq})
\end{equation}
where $\hat{\mG}_{\ox}()$ are the combined operator and exoskeleton
gravitational forces calculated using an imperfect model,

@@

Friction can also be compensated using feedforward control.
\begin{equation}
\vtau_{\myx} = \hat{\mF}_{\myx}(\hat{\vq},\hat{\dot{\vq}}) 
\end{equation}
are the estimated friction torques at the estimated exoskeleton configuration
and velocity.
Note that friction may be load dependent, and may require additional state to
correctly handle stiction and hysteresis effects.

@@

What is admittance?
\url{https://en.wikipedia.org/wiki/Admittance}
Compliance, inverse damping, and inverse mass are components of an admittance,
as they map force and torques to kinematic variables (position, velocity, and
acceleration).

We can implement operator force control 
if measurements of the operator contact forces with
the exoskeleton are available.
The exoskeleton moves as to create as little force on the operator as possible 
assuming the actuators are torque sources:
\begin{equation}
\vtau_{\myx} = \vtau_{\xinvdynw} - \mJ_{\ox}^{\tr}(\vq) {\vf_{\ox}}_d
- \mJ_{\ox}^{\tr}(\vq) \mK_{\vf_{\ox}} ( \hat{\vf}_{\ox} - {\vf_{\ox}}_d )
\end{equation}
$\mK_{\vf_{\ox}}$ is a gain matrix. $\mJ_{\ox}$ is the Jacobian matrix for the operator
force $\vf_{\ox}$.
Note that this does not require inverting the Jacobian matrix.
We will see that the position control version does invert the Jacobian matrix.

@@

So far we have ignored or canceled the effect of 
the exoskeleton contact with the outside world.
A different control objective could be to map the external forces to 
desired operator forces:
\begin{equation}
{\vf_{\ox}}_d = \mF_{\ox-\xw} ( \vf_{\xw} )
\end{equation}
and cancel the external forces not transferred in the inverse dynamics.
Everything in the above equation is estimated, so we are dropping the hats 
$\hat{\mbox{}}$.

@@

This is usually done by using stored trajectories or trajectories generated
online that specify desired position $\vq_d$, 
desired velocity $\dot{\vq}_d$, and
desired acceleration $\ddot{\vq}_d$.
The feedforward method is to compute the needed torques (potentially offline)
using inverse
dynamics (including the operator link inertias, link centers of mass, and link
moments of inertia in the dynamic model) based on the desired position, 
desired velocity, and desired acceleration. Trajectory tracking errors are compensated
for by feedback control designed using some other method. In most robotics
fixed gain independent joint linear control is used.
This method is safest in the sense that the feedback controller can be tested
and verified independently of the desired trajectories and dynamic models.

@@

Another control objective would be to allow the exoskeleton-world contact to
be compliant or have an assigned impedance.
Let's describe a small position offset of the exoskeleton-world
contact points as $\vq_w$. We know that:
\begin{equation}
\dot{\vq}_w = \mJ_{\xw}(\vq) \dot{\vq}
\end{equation}
and that
\begin{equation}
\vtau_{\myx} = \mJ_{\xw}^{\tr}(\vq) \vf_{\xw} 
\end{equation}
This allows us to relate stiffness and damping in the external world coordinates
with the stiffness and damping in exoskeleton joint coordinates.
Here is the relationship for damping:
\begin{equation}
\vtau_{\myx} = \mJ_{\xw}^{\tr}(\vq) \mK_{\dot{w}} \mJ_{\xw}(\vq) \dot{\vq}
\end{equation}
So the damping in joint coordinates is $\vtau_{\myx} = \mK_{\dot{\vq}} \dot{\vq}$ and
therefore
\begin{equation}
\mK_{\dot{\vq}}(\vq) = \mJ_{\xw}^{\tr}(\vq) \mK_{\dot{w}} \mJ_{\xw}(\vq)
\end{equation}
The mapping is the same for stiffness:
\begin{equation}
\mK_{\vq}(\vq) = \mJ_{\xw}^{\tr}(\vq) \mK_{w} \mJ_{\xw}(\vq)
\end{equation}
So we should assign the exoskeleton joint position gains to be $\mK_{\vq}(\vq)$ and
the joint velocity gains to be $\mK_{\dot{\vq}}(\vq)$ in equation~\ref{eq:impedance}.

@@

This section describes how to handle underactuation and overactuation using
``floating body dynamics.''
We have been sweeping this
issue under the rug so far. It is now useful to be more explicit.
There are two kinds of underactuation we want to consider. The first
is ``floating body dynamics'', where the robot is not bolted to the floor.
Contacts are made and broken during walking, running, and many other behaviors,
and contacts slip.
Humanoid robots (and humans) have floating body dynamics.
A ``root'' location (usually somewhere
on the pelvis) is chosen, and the position, velocity, orientation, and angular
velocity of that point and the rigid body it is on
are included in the state vector $\vq$. The corresponding dimensions of the
velocity vector $\dot{\vq}$ are not actuated. A Jacobian matrix
$\mJ_{\myx}$ is used to map from actuation variables $\vu$ to joint torques $\vtau$:
\begin{equation}
\vtau_{\myx} = \mJ_{\myx} \vu_{\myx}
\end{equation}
The rows in $\mJ_{\myx}$ corresponding to the root linear and angular velocities
are all zero.
To maintain compatibility in our joint representation, we will similarly
map operator ``muscle commands'' $\myo$ to operator joint torques $\vtau_{\myo}$.
\begin{equation}
\vtau_{\myo} = \mJ_{\myo} \vu_{\myo}
\end{equation}

The second kind of underactuation appears in exoskeletons when we rely on the
operator to actuate a particular joint. For example, many lower body
exoskeletons only actuate the joints in the sagittal plane, leaving lateral
and twist (yaw) joints unactuated or passively actuated. 
Many ``exoskeletons'' or orthoses only actuate one joint on each limb (typically
the ankle or knee). In these situations the question arises as to what is
the scope of control: should the device only consider
just the actuated joints,
or should it consider a wider range of operator joints? Many of these devices
do not have enough sensing (such as an IMU) to even estimate the state of other
operator joints. We believe that for high performance the entire state of the
operator must be considered in control, and we describe a way to do it in
Section~\ref{sec:qp}.
However, we are not aware of any system that goes this far in terms of control.

@@

``Missing joint''
underactuation can be handled by reducing the size of the $\vu_{\myx}$ vector
and changing the matrix $\mJ_{\myx}$ to reflect the missing actuation.
We can have redundant actuation (more than one actuator across any particular
joint). Here we augment $\vu_{\myx}$ and $\mJ_{\myx}$ to accommodate extra actuators.
We can have multi-joint actuation, which can be handled in $\vu_{\myx}$ 
and $\mJ_{\myx}$ as well.

@@

{\bf Open loop control} 
and
{\bf fixed feedforward control with feedback control}
continue to work with underactuation
as long as a drive function can be found
to generate the desired behavior.

@@

Full (operator) body
{\bf online inverse dynamics} as described in Section~\ref{sec:invdyn}
no longer works if $\mJ_{\myx}$ is not invertible (not square or not full rank).
In the underactuated and overactuated cases $\mJ_{\myx}$ is not square. 
For a partial (operator) body
exoskeleton we can rewrite the dynamics in terms of only those
joints the exoskeleton actuates, and invert those dynamics to cancel exoskeleton
effects on those joints. This is what has been done with the BLEEX exoskeleton,
for example.~\cite{}.

@@

Quadratic programming~\cite{Wikipedia} solves the following problem:
Minimize
\begin{equation}
\vx^{\tr} \mQ \vx + c^{\tr} \vx
\end{equation}
with inequality constraints
\begin{equation}
\mA_1 \vx \le \vb_1
\end{equation}
and equality constraints
\begin{equation}
\mA_2 \vx = \vb_2
\end{equation}
The optimization variable $\vx$ includes accelerations $\ddot{\vq}$, actuator
commands $\vu$, and contact forces $\vf$.
We could optimize only taking into account the exoskeleton and its dynamics,
minimizing the discrepancy between actual and desired accelerations, actuator command
magnitude, and contact force magnitude:
\begin{equation}
C_{\myx}(\vx) = (\vq - \vq_d)^{\tr} \mW_{\vq} (\vq - \vq_d)
+ \vu_{\myx}^{\tr} \mW_{\vu_{\myx}} \vu_{\myx}
+ \vf_{\ox}^{\tr} \mW_{\vf_{\ox}} \vf_{\ox}
+ \vf_{\xw}^{\tr} \mW_{\vf_{\xw}} \vf_{\xw}
\end{equation}
$\mW_{\vq}$, $\mW_{\vu_{\myx}}$, $\mW_{\vf_{\ox}}$,
and $\mW_{\vf_{\xw}}$ are weight matrices to trade off the
different errors/magnitudes and error/magnitude directions.
The dynamics (equation~\ref{eq:xdynamics2}) are enforced as an equality constraint.
Actuator limits, friction limits, center of pressure location constraints,
and unidirectional contact constraints are all
enforced as inequality constraints.
Variants of this optimization
were performed every 1-2ms by various teams to control the
Atlas robots in the DARPA Robotics Challenge.
